---
title: "Data Import"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Import}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE}
library(amerifluxr)
library(pander)
```

amerifluxr is a programmatic interface to the [AmeriFlux] (https://ameriflux.lbl.gov/). This vignette demonstrates examples to import data and metadata downloaded from AmeriFlux, and to parse and clean data for further use. A [companion vignette for site slection](site_selection.html) before download is available as well.

## Download data

Both data and metadata can be downloaded through the [AmeriFlux data portal] (https://ameriflux.lbl.gov/data/download-data/). Users will need to create a personal AmeriFlux account first. Once logged in, users can select a list of sites to download using the Site ID (CC-Sss). Check site_selection for site selection. The flux/met data products (aka. BASE data product) are organized by sites, with one zipped file (e.g., CC-Sss_BASE-BADM_#-#.zip) for each site.   

While downloading BASE data product, users can also choose to download additional metadata files (aka. BADM data product) for all AmeriFlux sites. If so, a single Excel file (e.g., AMF_AA-Flx_BIF_YYYYMMDD.xlsx, AMF_AA-Net_BIF_YYYYMMDD.xlsx) will be provided in the download.

For this vignette, we will use following example data files [files are truncated to limit package size, for demonstration purpopses only].

```{r echo=TRUE}
# An example of BASE zipped files downloaded for US-CRT site
system.file("extdata", "AMF_US-CRT_BASE-BADM_2-5.zip", package = "amerifluxr")

# An example of unzipped BASE files from the above zipped file
system.file("extdata", "AMF_US-CRT_BASE_HH_2-5.csv", package = "amerifluxr")

# An example of all sites' BADM data
system.file("extdata", "AMF_AA-Flx_BIF_20201218.xlsx", package = "amerifluxr")

```

# BASE data product
## Import BASE data product

The amd_read_base() can be used to import BASE file, either from a zipped file or an unzipped comma-separated file. The parse_timestamp parameter can be used if additional time-keeping columns (e.g., year, month, day, hour) are desired. 

```{r results = "asis"}
# read the BASE from a zip file, without additional parsed time-keeping columns
base1 <- amf_read_base(
  file = system.file("extdata",
                     "AMF_US-CRT_BASE-BADM_2-5.zip",
                     package = "amerifluxr"),
  unzip = TRUE,
  parse_timestamp = FALSE
)
pander::pandoc.table(base1[c(1:3),])

# read the BASE from a csv file, with additional parsed time-keeping columns
base2 <- amf_read_base(
  file = system.file("extdata",
                     "AMF_US-CRT_BASE_HH_2-5.csv",
                     package = "amerifluxr"),
  unzip = FALSE,
  parse_timestamp = TRUE
)
pander::pandoc.table(base2[c(1:3), c(1:10)])
```

## Parse and interpret BASE data product

The detailed explanation of BASE data format and variables can be found on [AmeriFlux website] (https://ameriflux.lbl.gov/data/aboutdata/data-variables/). In short, TIMESTAMP_START and TIMESTAMP_END (i.e., YYYYMMDDHHMM 12 digits) refer to the starting and ending time of each reporting interval (i.e., rows). All other variables use the format of <base name>_<qualifier>, e.g., FC_1, CO2_1_1_1. Base names indicate fundamental quantities that are either measured or calculated / derived. Qualifiers are suffixes appended to variable base names that provide additional information (e.g., gap-filling, position) about the variable. In some cases, qualifiers are omitted if only one variable is provided for a site.

The amf_variable() can be used to retrieve the latest list of base names and default units. For sites that don't have many variables and less-complicated qualifiers, the users should be able to interpret the variables and their qualifiers. The amf_variable() also returns the expected maximal and minimal values based on physically plausible ranges or network-wide  reported values.   

```{r results = "asis"}
# get a list of latest base names and units. 
FP_ls <- amf_variables()
pander::pandoc.table(FP_ls[c(1:10), ])
```

Alternatively, the amf_parse_basename() can be use to programmatically parse the the variable names into base names and qualifiers. This function can be helpful for sites with many and often complicated qualifiers, as a prerequisite for handling data from many sites. The function returns a data frame with information about each variable's base name and qualifier, whether a variable is gap-filled, layer-aggregated, or replicate aggregated.

```{r results = "asis"}
# parse the variable name
basename_decode <- amf_parse_basename(var_name = colnames(base1))
pander::pandoc.table(basename_decode[c(1, 2, 3, 4, 6, 11, 12),])
```

## Data filtering

While all BASE data products are quality-checked before release, the data may not be filtered for all outliers. The amf_filter_base() can be use to filter the data based on the expected physically ranges (i.e., obtained through amf_variables()). By default, a ±5% buffer is applied to account for possible edge values near the lower and upper bounds, which are commonly observed for radiation variables, relative humidity, and snow depth. 

```{r}
# filter data, using default physical range +/- 5% buffer
base_f <- amf_filter_base(data_in = base1)

```

## Measurement height information

Measurement Height information contains height/depth and instrument model information for AmeriFlux BASE data products. The info can be downloaded directly using the amf_var_info() function. The function returns a data frame for all available sites, and can subset using the "Site_ID" column. The "Height" column refers to the distance from the ground surface in meters. Positive values are heights, and negative values are depths. See the [web page](https://ameriflux.lbl.gov/data/measurement-height/) for explanation. 

```{r results = "asis"}
# obtain the latest measurement height information
var_info <- amf_var_info()

# subset the variable by target Site ID
var_info <- var_info[var_info$Site_ID == "US-CRT", ]
pander::pandoc.table(var_info[c(1:10), ])
```

# BADM data product
## Import BADM data

Biological, Ancillary, Disturbance, and Metadata (BADM) are non-continuous information that describe and complement continuous flux and meteorological (Flux/Met) data. BADM include general site description, metadata about the sensors and their setup, maintenance and disturbance events, and biological and ecological data that characterize a site’s ecosystem. See [link](https://ameriflux.lbl.gov/data/badm/badm-basics/) for details.

The amf_read_bif() can be used to import the BADM data file. The function returns a data frame for all available sites, and can subset using the "SITE_ID" column. 

```{r results = "asis"}
# read the BADM BIF file, using an example data file
bif <- amf_read_bif(file = system.file("extdata",
                                      "AMF_AA-Flx_BIF_20201218.xlsx",
                                       package = "amerifluxr"))

# subset by target Site ID
bif <- bif[bif$SITE_ID == "US-CRT", ]
pander::pandoc.table(bif[c(1:15), ])

# get a list of all BADM variable groups and variables
unique(bif$VARIABLE_GROUP)
length(unique(bif$VARIABLE))

```

As shown above, BADM data contain information from a variety of variable groups. Browse the available variable groups [here](https://ameriflux.lbl.gov/data/badm/badm-standards/).  

To get the BADM data for a certain variable group, use amf_extract_badm() function.

```{r results = "asis"}
# extract the FLUX_MEASUREMENTS group
bif_flux <- amf_extract_badm(bif_data = bif, select_group = "GRP_FLUX_MEASUREMENTS")
pander::pandoc.table(bif_flux)

# extract the HEIGHTC (canopy height) group
bif_hc <- amf_extract_badm(bif_data = bif, select_group = "GRP_HEIGHTC")
pander::pandoc.table(bif_hc)
```

Note: amf_extract_badm() returns all columns in characters. Certain groups of BADM variables contain columns of time stamps (i.e., ISO format) and data values, and need to be converted before further use.     

```{r fig.width = 7}
# convert HEIGHTC_DATE to POSIXlt
bif_hc$TIMESTAMP <- strptime(bif_hc$HEIGHTC_DATE, format = "%Y%m%d", tz = "GMT")

# convert HEIGHTC column to numeric
bif_hc$HEIGHTC <- as.numeric(bif_hc$HEIGHTC)

# plot time series of canopy height
plot(bif_hc$TIMESTAMP, bif_hc$HEIGHTC, xlab = "TIMESTAMP", ylab = "canopy height (m)")

```

Also useful, the contacts of the site members and data DOI can be obtained from the BADM data.

```{r results = "asis"}
# get a list of contacts
bif_contact <- amf_extract_badm(bif_data = bif, select_group = "GRP_TEAM_MEMBER")
pander::pandoc.table(bif_contact)

# get data DOI
bif_doi <- amf_extract_badm(bif_data = bif, select_group = "GRP_DOI")
pander::pandoc.table(bif_doi)
```

